Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-10-29 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Reviewed intro with Christine

The use of Shannon entropy $-\sum p_i \log p_i$ is well established in very different fields of knowledge such as computer science[cite], physics[cite] and ecology[cite]. It is a fundamental element in information theory since Shannon developed his mathematical theory of communication. It has been used extensively within statistical mechanics to link the microscopic description of a system to its macroscopic thermodynamic one. It has been used in ecology to characterize biodiversity and calculate species distributions under given constraints. Entropy maximization has been shown to be a powerful tool in all these settings.

While the techniques are well developed, it is puzzling that this concept can be applied in such different contexts, given that it assumes different features in each one. For example, in information theory the Shannon entropy for a book will change whether we choose to encode words or letters, which means the quantity does not have a fully objective feature: it depends on an arbitrary choice. This is not problematic for an information theoretic quantity, but it is for a physical quantity. In statistical mechanics the $p_i$ represent probabilities, which means the Shannon entropy is a property of an ensemble of physical systems. This leads some to claim that the Shannon entropy is epistemic in nature and represents ignorance. [Add references] In ecology, however, the $p_i$ represent the fraction of individuals belonging to the $i$th species. There is no probabilistic ensemble to which one can associate a degree of ignorance. Finally, thermodynamic entropy is associated to the notion of states and irreversible time evolution, which are not well defined concepts for the other fields. The same expression, then, seems to have acquire different flavors and it is not clear why a single concept would apply in all these cases.

To make matter worse, it is fairly typical to describe entropy with notions such as uncertainty[cite], knowledge[cite], lack of knowledge[cite], disorder[cite]. While some of these characterizations are useful in some contexts to give an intuitive picture, they do not seem to always work. A common objection to the characterization as disorder, for example, is that a uniform distribution corresponds to a homogeneous system, which one would expect to be maximally ordered; yet the distribution maximizes the Shannon entropy, which would indicate maximal disorder. Moreover, these notions remain vague in the sense that the formula is not formally derived by those concepts and therefore it is not clear why that particular indicator would be an appropriate choice. We are left to wonder: is there really some single underlying concept, that, when applied appropriately in the different fields, reduces to the correct notion of entropy? Or is it just a big coincidence?

We propose that there is indeed a characterization of Shannon's formula that is both more general and more appropriate; that is independent on whether it is applied to sources of messages, ensembles of microstates or animal populations. We will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements}. That is, given a distribution of some elements over some quantity, the Shannon entropy represents how varied, how different from each other, the values are. In fact, it is the only continuous and linear index that measures the degree of variability within a distribution, and those requirements lead naturally to same three conditions that Shannon set for his formula. In other words, the concept of variability with a distribution leads in a formal way to the correct expression. We will then show how this perspective naturally connects to information theory and statistical mechanics when applied in the respective contexts.

As we believe this is the right characterization, we will therefore use the expression Shannon ``variability'', rather than ``entropy'' or ``information'', as it tells us exactly what the quantity describes. It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Boltzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. We will reserve the word entropy for the thermodynamic entropy, the one dealing with physical states and irreversible processes.

This characterization also explains why the Shannon variability has such a broad range of applicability. As nature presents itself in a multitude of forms and combinations, a fundamental task in science is to recognize which objects belong together and study how their different attributes are expressed in different ways or degrees. They may galaxies, animals or outcomes of the same process; we may study their shape, behavior or numerical quantities. But the idea of grouping them and try to understand the source of their variation is a common.\footnote{In fact, determining which collection of objects is significant and what attributes are relevant is significant part of the work.} Measuring and explaining this variability is an essential task. It is this very variety that is the inspiration for and the subject of scientific investigation. A universe without variety, where all objects were the same and remained the same, is barely conceivable and certainly it would leave no space for scientific investigation.

As a final note, we want to stress that the goal of this work is not to present some new technical result. The goal is to put all details together in a coherent picture that gives them context and see precisely how they are related. This can help, especially the new practitioners, to avoid common misconceptions. Moreover, while some details presented here may be well known in a particular community, we find that it may be new to another. For example, we will show an example of Huffmann coding to illustrate the meaning of the numerical value associated with the Shannon variability. This is standard practice in computer science courses, yet it is not in statistical mechanics. Therefore restating all the crucial aspects in one place is beneficial because the connection between them is not always readily available.

@@

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the probability. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ will be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity linear when combining two independent distributions.

>>>>

The use of Shannon entropy $-\sum p_i \log p_i$ is well established in very different fields of knowledge such as computer science[cite], physics[cite] and ecology[cite]. Within computer science it is a fundamental element in information theory since Shannon developed his mathematical theory of communication. It has been used extensively within statistical mechanics to link the microscopic description of a system to its macroscopic thermodynamic one. It has been used in ecology to characterize biodiversity and calculate species distributions under given constraints. Entropy maximization has been shown to be a powerful tool.

While the techniques are well developed, it is puzzling that this concept can be applied in such different contexts, given that it assumes different features in each one. For example, in information theory the Shannon entropy for a book will change whether we choose to encode words or letters, which means the quantity does not have a fully objective feature: it depends on an arbitrary choice. This is not problematic for an information theoretic quantity, but it is for a physical quantity. In statistical mechanics the $p_i$ represent probabilities, which means the Shannon entropy is a property of an ensemble of physical systems. This leads some to claim that the Shannon entropy is epistemic in nature and represents ignorance. [Add references] In ecology, however, the $p_i$ represent the fraction of individuals belonging to the $i$th species. There is no probabilistic ensemble to which one can associate a degree of ignorance. Finally, thermodynamic entropy is associated to the notion of states and irreversible time evolution, which are not well defined concepts for the other fields. The same expression, then, seems to have different traits and it is not clear why a single concept would apply in all these cases.

To make matters worse, it is fairly typical to describe entropy with notions such as uncertainty[cite], knowledge[cite], lack of knowledge[cite], disorder[cite]. While some of these characterizations are useful in some contexts to give an intuitive picture, they do not always seem to work. A common objection to the characterization as disorder, for example, is that a uniform distribution corresponds to a homogeneous system, which one would expect to be maximally ordered; yet the distribution maximizes the Shannon entropy, which would indicate maximal disorder. Moreover, these notions remain vague in the sense that the formula is not formally derived from those concepts, and therefore it is not clear why that particular indicator would be an appropriate choice. We are left to wonder: is there really some single underlying concept, that, when applied appropriately in the different fields, reduces to the correct notion of entropy? Or is it just a big coincidence?

We propose that there is indeed a characterization of Shannon's formula that is both more general and more appropriate; that is independent of whether it is applied to sources of messages, ensembles of microstates or animal populations. We will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements}. That is, given a distribution of some elements over some quantity, the Shannon entropy represents how varied, how different from each other, the values are. In fact, it is the only continuous and linear index that measures the degree of variability within a distribution, and those requirements lead naturally to the same three conditions that Shannon set for his formula.\cite{Shannon} In other words, the concept of variability within a distribution leads in a formal way to the correct expression. In all cases where the formula is applied, in fact, we are using distributions as objects of our investigation and the variability within that distribution will be of interest. Yet, the physical meaning of those distributions can be widely different: they may represent the fractional distribution of animals according to their species; they may represent probability distributions of different outcomes; they may represent knowledge of certain facts as in Bayesian statistics. Therefore the variability itself, in the context of these different cases, will necessarily acquire a different meaning. In particular, we will see how this perspective naturally unfolds in the contexts of information theory and statistical mechanics.

As we believe this is the right characterization, we will therefore use the expression Shannon ``variability'', rather than ``entropy'', ``information'' or ``diversity index'', as it tells us exactly what the quantity describes. It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Boltzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. We will reserve the word ``entropy'' for the thermodynamic entropy, the one dealing with physical states and irreversible processes.

This characterization also explains why the Shannon variability has such a broad range of applicability. As nature presents itself in a multitude of forms and combinations, a fundamental task in science is to recognize which objects belong together and study how their different attributes are expressed in different ways or degrees. They may be galaxies, animals or outcomes of the same process; we may study their shape, behavior or numerical quantities. But the idea of grouping them and trying to understand the source of their variation is a common endeavor in science.\footnote{In fact, determining which collection of objects is significant and what attributes are relevant is a significant part of the work.} Measuring and explaining this variability is an essential task. It is this very variety that is the inspiration for and the subject of scientific investigation. A universe without variety, where all objects were the same and remained the same, is barely conceivable and certainly it would leave no space for scientific investigation. In the words of Leibniz, the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.\footnote{MOVE TO BIBLIO Paragraph 58 of Leibniz's \emph{Mondadology}}

As a final note, we want to stress that the goal of this work is not to present some new technical result. The goal is instead to synthesize all details into a coherent picture that gives them context and a precise relationship to each other. This can help, especially the new practitioners, to avoid common misconceptions. Moreover, while some details presented here may be well known in a particular community, we find that they may be new to another. For instance, we will show an example of Huffmann coding to illustrate the meaning of the numerical value associated with the Shannon variability. This is standard practice in computer science courses, yet it is not in statistical mechanics. Therefore restating all the crucial aspects in one place is beneficial to make the connection between them readily available.

@@

%https://www.sciencedirect.com/science/article/abs/pii/S0169534714001037

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the probability. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ to be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity additive when combining two independent distributions.
